Soundness Proofs for Iterative Deepening

### 2016-09-01 oopsla-16/proof/assumptions.tex
Import proof
oopsla proofs: fix discussion of which PPs happen before and/or after associated instructions

{\bf No forcibly blocking other threads.}
We assume the sync API behaves much like pthreads and/or message-passing:
that is, a thread can cause itself to block on some condition,
and other threads may wake it up,
but there exists no primitive which one thread can use to revoke another thread's runnable status when that other thread is not itself using the sync API concurrently.

{\bf Halting.}
From a certain perspective, proving convergence is the same as proving completeness,
which should be impossible for any runtime analysis of a Turing-complete language \cite{entscheidungsproblem}.
However, being already limited by practical real-world CPU budgets,
we are already accepting that many tests will time out.
We are concerned with proving the limited case that when \quicksand~{\em does} terminate with all state spaces completed, the verification is sound.
Inversely, during our proof, when we show that Iterative Deepening will eventually converge to an arbitrary buggy interleaving, we assume that no intermediate state spaces contain nontermination bugs.
If they do, for the sake of testing/verification, we are satisfied with finding that bug instead.
In this proof, we assume that the MC's heuristic infinite loop checker has no false negatives;
i.e., it will never get stuck forever in an infinite loop without identifying a bug, while occasionally producing false positives.
% , such as for a program trying to find a counterexample for Sally's Conjecture.
% TODO: cite some unproved mathematical conjecture by a woman, as an example of an undecidable program?

>>>>

%{\bf No forcibly blocking other threads.}
%We assume the sync API behaves much like pthreads and/or message-passing:
%that is, a thread can cause itself to block on some condition,
%and other threads may wake it up,
%but there exists no primitive which one thread can use to revoke another thread's runnable status when that other thread is not itself using the sync API concurrently.
%
%We require this because of the asymmetry between synchronization PPs (which occur {\em after}) synchronization operations)
%and data-race PPs (which occur {\em before} their associated read/write).
%For example, if a primitive $\mathsf{deschedule\_other}$ existed,
%a transition such as $\mathsf{write}(a,v); \mathsf{deschedule\_other}(t_2)$ would lack a PP between the two instructions,
%which could be necessary to expose some...

{\bf Halting.}
From a certain perspective, proving convergence is the same as proving completeness,
which should be impossible for any runtime analysis of a Turing-complete language \cite{entscheidungsproblem}.
However, being already limited by practical real-world CPU budgets,
we are already accepting that many tests will time out.
We are concerned with proving the limited case that when \quicksand~{\em does} terminate with all state spaces completed, the verification is sound.
Inversely, during our proof, when we show that Iterative Deepening will eventually converge to an arbitrary buggy interleaving, we assume that no intermediate state spaces contain nontermination bugs.
If they do, for the sake of testing/verification, we are satisfied with finding that bug instead.
In this proof, we assume that the MC's heuristic infinite loop checker has no false negatives;
i.e., it will never get stuck forever in an infinite loop without identifying a bug, while occasionally producing false positives.

### 2016-09-01 oopsla-16/proof/di.tex
oopsla proofs: fix discussion of which PPs happen before and/or after associated instructions

We use {\em synchronization API PPs} to denote the set of predicates
%statically-available
which occur immediately after any of $\mathcal{A}_{sync}$.
%$\mathsf{mutex\_lock}$,
%$\mathsf{mutex\_unlock}$,
%$\mathsf{deschedule}$,
%$\mathsf{make\_runnable}$,
%$\mathsf{thread\_fork}$,
%$\mathsf{thread\_exit}$,
%$\mathsf{yield}$.
Because no other instruction affects a thread's runnability, it is always possible to execute a program by switching the currently-executing thread only at synchronization API PPs.

All data-race PP predicates will occur immediately {\em before} a $\mathsf{read}$, $\mathsf{write}$, or atomic swap.
Other predicates are possible, though we will show they are irrelevant.

@@

Intuitively, MHB expresses the "cannot be reordered with" (or "enables") relation between transitions.
Two transitions $A$ and $B$ of different threads MHB if some synchronization event in $A$ causes $B$ to become runnable while it was previously blocked. Such synchronization events include {\tt thread\_create}, {\tt cond\_signal}, {\tt sem\_post}, but {\em not} {\tt mutex\_lock} or {\tt mutex\_unlock}.

@@

Intuitively, the behaviour of a program could change by reordering two transitions only if they contain a memory conflict.
All possible interleavings of a program can be partitioned into equivalence classes,
so only one interleaving from each equivalence class need be tested to ensure total schedule coverage \cite{mazurkiewicz}.

@@

{\bf Intuition for Iterative Deepening convergence.}
In summary, we are proving that when Iterative Deepening finishes saturating the set of available data-race PPs,
that set represents every single code location where a preemption could possibly affect the program's behaviour,
and that completing the associated state spaces is as strong of a verification as testing all possible thread interleavings under any preemptions anywhere.
Some data-races may be hidden in control-flow paths which could only be executed after preempting on a different data-race,
but the technique's iterative nature will eventually find it.
On the other hand, relying on the soundness of DPOR, preempting on an instruction which is neither a data-race or sync API boundary cannot affect the program's behaviour,
so any PPs beyond the ones we consider are irrelevant.

>>>>

We use {\em synchronization API PPs} to denote the set of predicates
%statically-available
which occur immediately before or after any of $\mathcal{A}_{sync}$.
%$\mathsf{mutex\_lock}$,
%$\mathsf{mutex\_unlock}$,
%$\mathsf{deschedule}$,
%$\mathsf{make\_runnable}$,
%$\mathsf{thread\_fork}$,
%$\mathsf{thread\_exit}$,
%$\mathsf{yield}$.
Because no other instruction affects a thread's runnability, it is always possible to execute a program by switching the currently-executing thread only at synchronization API PPs.

All data-race PP predicates will occur immediately before a $\mathsf{read}$, $\mathsf{write}$, or atomic swap.
Other predicates are possible, though we will show they are irrelevant.

@@

Intuitively, MHB expresses the "cannot be reordered with" relation between transitions
("enables" in DPOR terminology \cite{dpor}).
Two transitions $A$ and $B$ of different threads MHB if some synchronization event in $A$ causes $B$ to become runnable while it was previously blocked.
Such synchronization events include $\mathsf{thread\_fork}$, $\mathsf{make\_runnable}$,
and sometimes but not always $\mathsf{mutex\_unlock}$.

@@

Intuitively, the behaviour of a program could change by reordering two transitions only if they contain a memory conflict.
All possible interleavings of a program can be partitioned into equivalence classes,
so only one interleaving from each equivalence class need be tested to ensure total schedule coverage \cite{mazurkiewicz}.
Equivalence is, of course, transitive.

@@

{\bf Intuition for Iterative Deepening convergence.}
In summary, we are proving that when Iterative Deepening saturates the set of data-race PPs,
that set represents every single instruction where a preemption could possibly affect the program's behaviour,
and that completing the associated state spaces is as strong of a verification as testing all possible thread interleavings by preempting anywhere.
A data-race may be hidden in control-flow paths reachable only after preempting on a different data-race,
but the technique's iterative nature will eventually find it.
On the other hand, relying on the soundness of DPOR, preempting on an instruction which is neither a data-race or sync API boundary cannot affect the program's behaviour,
so any such PPs are irrelevant.

Note that while we defined synchronization API PPs to occur both before and after any $\mathcal{A}_{sync}$ instruction,
we will only add data-race PPs {\em before} their associated $\mathsf{read}$ or $\mathsf{write}$.
Our proof requires preempting only before each racing instruction, not after,
in order to fully saturate the PP set with all possible data races.
Then, once saturated, preempting after each racing instruction would be extraneous,
because all subsequent instructions before the next PP must be local operations or non-communicating reads/writes%.
% Adding a PP both before and after could potentially expose some bugs sooner,
% if the 'after' PP were redundant with another data-race PP but that other race would take forever to find.
% So, I think it's an optimization to do it this way (although I haven't done a comparison experiment).
%
\footnote{Our implementation also optimizes the synchronization API PPs,
generally preempting only {\em after} each synchronization instruction.
$\mathsf{mutex\_lock}$ is the exception, as it can cause other threads to become blocked.
All the others can only make blocked threads runnable, establishing MHB,
which also provides MHB for the preceding transition of the invoking thread.
Hence, coalescing those transitions does not exclude any possible interleavings.
% If a third thread could "query" whether the 2nd thread is still blocked (or nonexistent),
% that would break the soundness of this optimization, and you'd need to preempt before mx_unlock and thr_create after all.
% XXX: In the implementation, such queries are actually possible with m_r or yield;
% while they do not interact with mutex_unlock, I think an adversary could use them to interleave with thr_create/exit.
}.
